Live Functional Programming with Typed Holes

### 2018-04-01 calculus.tex
finished draft of 3.2

We will now make the intuitions developed in the previous section formally precise by specifying the \HazelnutLive core calculus and  its accompanying metatheory. We have mechanized these formal developments using the Agda proof assistant \cite{norell:thesis,norell2009dependently}. Sec.~\ref{sec:agda-mechanization} provides some additional details on this mechanization and the supplemental material includes the full mechanization. % The proofs of the metatheorems are established by this mechanization; the proofs are only very briefly outlined in this section.

The syntax of the core calculus, specified in Fig.~\ref{fig:hazelnut-live-syntax}, consists of types and expressions with holes. We distinguish between \emph{external} expressions, $e$, and \emph{internal} expressions, $d$. External expressions correspond to programs as entered by the programmer (see Sec.~\ref{sec:implementation}\todo{or maybe Sec. 2?}{} for discussion of manual, semi-automated and fully automated hole entry methods). Each well-typed external expression (as specified in Sec.~\ref{sec:external-statics} below) expands to a well-typed internal expression (Sec.~\ref{sec:expansion}) before it is evaluated (Sec.~\ref{sec:evaluation}). We distinguish the external and internal languages because (1) the external language supports type inference and explicit type ascriptions, $\hexp : \htau$, but it is  formally simpler to eliminate ascriptions and specify a type assignment system when defining the dynamic semantics\todo{cite frank's notes}; and (2) we need additional syntactic machinery during evaluation for tracking hole closures and dynamic type casts. This machinery is inserted by the expansion step, rather than entered explicitly by the programmer. In this regard, the internal language is analagous to the cast calculus in the gradually typed lambda calculus \cite{DBLP:conf/snapl/SiekVCB15,Siek06a}, though as we will see the \HazelnutLive internal language goes beyond the cast calculus in several respects.

@@

Let us start with a brief overview of the static semantics of the \HazelnutLive external language. Types, $\htau$, classify external expressions, $e$, according to the type system in Fig. \ref{fig:bidirectional-typing}, which closely follows the \Hazelnut type system \cite{popl-paper} (the minor differences are discussed as they come up). The type system is specified in the \emph{bidirectional} style \cite{Pierce:2000ve,bidi-tutorial,DBLP:conf/icfp/DunfieldK13,Chlipala:2005da} around two mutually defined judgements. The type synthesis judgement, $\hsyn{\hGamma}{\hexp}{\htau}$, synthesizes a type $\htau$ for $\hexp$ assuming $\hGamma$, which tracks typing assumptions of the form $x : \htau$ in the usual manner \cite{pfpl,tapl}. The type analysis judgement, $\hana{\hGamma}{\hexp}{\htau}$, instead checks $\hexp$ against a given $\htau$. Algorithmically, the type is an input of type analysis.% Algorithmically, the type is an output of type synthesis but an input of type analysis.

The benefit of specifying the \HazelnutLive external language bidirectionally is that the programmer does not need to annotate each hole with a type. Empty holes are written $\hehole{u}$, where $u$ is a hole name assumed unique (this notation is taken from \Hazelnut; hole names are new to \HazelnutLive). Rule {SEHole}\todo{rule name macros}{} specifies that empty holes synthesize hole type, written $\tehole$. If an empty hole appears where an expression of some other type is expected, e.g. under an explicit ascription $e : \htau$ (governed by Rule {SAsc}) or in the argument position of a function application (governed by Rule {SAp}, discussed below), we apply the \emph{subsumption rule}, Rule {ASubsume}, which specifies that if an expression $e$ synthesizes type $\htau$, then it can be checked against any \emph{consistent} type, $\htau'$. Fig.~\ref{fig:tconsistent} specifies the type consistency relation, written $\tconsistent{\htau}{\htau'}$, which coincides with the type consistency relation from gradual type theory by identifying the hole type with the unknown type~\cite{Siek06a}. Two types are consistent if they differ only up to type holes in corresponding positions. Note that type consistency is reflexive and symmetric but it is not transitive (quite unlike subtyping, which is anti-symmetric and transitive; subtyping can be integrated into a gradual type system following \citet{Siek:2007qy}). The hole type is consistent with every type, and so, by the subsumption rule, empty expression holes can appear where an expression of any type is expected.

@@

The rule governing function application, Rule {SAp}, similarly treats expressions of hole type in function position as if they were of type $\tarr{\tehole}{\tehole}$ using the matched arrow type judgement.

@@

Well-typed external expressions expand to well-typed internal expressions, $d$, for evaluation. The rules governing expansion are given in Fig.~\ref{fig:expansion} and the rules governing type assignment for internal expressions are given in Fig.~\ref{fig:hasType}.

Expansion is specified, much like the type system for the external language, in the bidirectional style by two mutually defined judgements. The synthetic expansion judgement, $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$, synthesizes a type, $\htau$, from $\hexp$, and produces an expansion, $d$, and a hole context, $\hDelta$. We say more about hole contexts below. The analytic expansion judgement, $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$, checks $\hexp$ against $\htau$ and produces an expansion, $d$, of type $\htau'$, and a hole context, $\hDelta$. The governing theorem below establishes that the type $\htau'$ is necessarily consistent with $\htau$.
\begin{thm}[Typed Expansion]\label{thm:typed-expansion} ~
  \begin{enumerate}[nolistsep]
    \item
      If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      then $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
    \item
      If $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
      then $\tconsistent{\htau}{\htau'}$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$.
  \end{enumerate}
\end{thm}
\noindent
The reason analytic expansion produces an expansion of consistent type is because the subsumption rule, as previously discussed, allows us to check an external expression against any type consistent with the type the expression actually synthesizes, while every internal expression can be assigned at most one type, i.e. the following standard unicity property holds of the type assignment system: 
\begin{thm}[Type Assignment Unicity]
  If $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$
  and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$
  then $\htau=\htau'$.
\end{thm}
\noindent
Consequently, analytic expansion reports the type actually assigned to the expansion it produces. For example, we can derive that $\expandAna{\hGamma}{c}{\tehole}{c}{b}{\emptyset}$, where $\emptyset$ is the empty hole context.

@@

The hole expansion rules are the only rules that introduce hypotheses, of the form $\Dbinding{u}{\hGamma}{\htau}$, into the hole context, $\Delta$. The purpose of the hole context is to record a typing context, $\Gamma$, and a type, $\tau$, for each hole name, $u$.\footnote{We use a hole context, rather than recording the typing context and type directly on each hole closure, to ensure that all closures for a hole name have the same typing context and type.} This notation for hole contexts is taken from contextual modal type theory (CMTT) \cite{Nanevski2008}, identifying hole names with metavariables and hole contexts with metavariable contexts (we say more about the connection with CMTT below). In all four hole expansion rules, the typing context recorded in the hole context is simply the current typing context when the hole is expanded. In the synthetic hole expansion rules, {ESEHole} and {ESNEHole}, the generated hole context assigns the hole type, $\tehole$, to $u$, as in the typing rules. However, the first two premises of the expansion subsumption rule, Rule EASubsume, disallow the use of subsumption for holes in analytic position during expansion. Instead, we have separate analytic rules, {EAEHole} and {EANEHole}, which record the type that the hole is being checked against into the hole context. This is again so that we can use a type assignment system for the internal language --- the type assignment rules TAEHole and TANEHole in Fig.~\ref{fig:hasType} assign the corresponding type from the hole context to the hole closure.

@@

\subsubsection{Cast Insertion} 
Consider the following example: $\hap{(\halam{x}{\tehole}{\hap{x}{c}})}{c}$. The type synthesized for this example viewed as an external expression is $\tehole$, because the hole type annotation on $x$ allows us to apply it as a function of type $\tarr{\tehole}{\tehole}$ as previously discussed, and the outer argument, $c$, can be checked against type $\tehole$ by subsumption. However, viewed as an internal expression, this example is not well-typed---we do not have subsumption in the type assignment system defined in Fig.~\ref{fig:hasType}. Indeed, it would violate type safety if we could assign a type to this example in the internal language, because beta reduction of this example viewed as an internal expression would result in $c(c)$, which is not well-typed ($c$ does not have arrow type). The problem is that leaving the argument type unknown leaves how the argument is being used (in this case, as a function) also unknown.\footnote{In a system where type reconstruction is first used to try to fill in type holes, we could express a similar example by using $x$ at two or more different types, thereby causing type reconstruction to fail. Rather than arbitrarily choosing one of the possible types for $x$ and inserting non-empty holes around uses that are inconsistent with that choice, we could leave the type hole unfilled and rely on the cast insertion machinery described here. On the other hand, if it is acceptable to arbitrarily choose one of the possible types, and type reconstruction is complete, then type holes will never appear in the internal language and the cast insertion machinery described in this section can be omitted entirely, leaving only the hole closure machinery described previously.}\todo{this footnote should probably be integrated into and refer to Sec. 2?}{} Recall that we can interpret hole types as unknown types from gradual type theory. This provides us with the solution to this problem: cast insertion.

@@

\input{fig-dynamics-aux}
%% \input{fig-dynamics-steps}
\input{fig-dynamics-contexts}

>>>>

We will now make the intuitions developed in the previous section formally precise by specifying the \HazelnutLive core calculus and  its accompanying metatheory. We have mechanized these formal developments using the Agda proof assistant \cite{norell:thesis,norell2009dependently} (see Sec.~\ref{sec:agda-mechanization} for some additional details and the supplemental material for the full mechanization). % The proofs of the metatheorems are established by this mechanization; the proofs are only very briefly outlined in this section.

The syntax of the core calculus, specified in Fig.~\ref{fig:hazelnut-live-syntax}, consists of types and expressions with holes. We distinguish between \emph{external} expressions, $e$, and \emph{internal} expressions, $d$. External expressions correspond to programs as entered by the programmer (see Sec.~\ref{sec:implementation}\todo{or maybe Sec. 2?}{} for discussion of manual, semi-automated and fully automated hole entry methods). Each well-typed external expression (as specified in Sec.~\ref{sec:external-statics} below) expands to a well-typed internal expression (see Sec.~\ref{sec:expansion}) before it is evaluated (see Sec.~\ref{sec:evaluation}). We distinguish the external and internal languages because (1) the external language supports type inference and explicit type ascriptions, $\hexp : \htau$, but it is  formally simpler to eliminate ascriptions and specify a type assignment system when defining the dynamic semantics\todo{cite frank's notes}; and (2) we need additional syntactic machinery during evaluation for tracking hole closures and dynamic type casts. This machinery is inserted by the expansion step, rather than entered explicitly by the programmer. In this regard, the internal language is analagous to the cast calculus in the gradually typed lambda calculus \cite{DBLP:conf/snapl/SiekVCB15,Siek06a}, though as we will see the \HazelnutLive internal language goes beyond the cast calculus in several respects.

@@

Let us start with an overview of the static semantics of the \HazelnutLive external language. Types, $\htau$, classify external expressions, $e$, according to the type system in Fig. \ref{fig:bidirectional-typing}, which closely follows the \Hazelnut type system \cite{popl-paper} (the minor differences are discussed as they come up below). The type system is specified in the \emph{bidirectional} style \cite{Pierce:2000ve,bidi-tutorial,DBLP:conf/icfp/DunfieldK13,Chlipala:2005da} around two mutually defined judgements. The type synthesis judgement, $\hsyn{\hGamma}{\hexp}{\htau}$, synthesizes a type $\htau$ for $\hexp$ assuming $\hGamma$, which tracks typing assumptions of the form $x : \htau$ in the usual manner \cite{pfpl,tapl}. The type analysis judgement, $\hana{\hGamma}{\hexp}{\htau}$, instead checks $\hexp$ against a given $\htau$. Algorithmically, the type is an output of type synthesis but an input of type analysis. At the top level, we start with synthesis. % Algorithmically, the type is an output of type synthesis but an input of type analysis.

The benefit of specifying the \HazelnutLive external language bidirectionally is that the programmer does not need to annotate each hole with a type. Empty holes are written $\hehole{u}$, where $u$ is a hole name assumed unique (hole notation is taken from \Hazelnut; hole names are new to \HazelnutLive). Rule {SEHole}\todo{rule name macros}{} specifies that empty holes synthesize hole type, written $\tehole$. If an empty hole appears where an expression of some other type is expected, e.g. under an explicit ascription (governed by Rule {SAsc}) or in the argument position of a function application (governed by Rule {SAp}, discussed below), we apply the \emph{subsumption rule}, Rule {ASubsume}, which specifies that if an expression $e$ synthesizes type $\htau$, then it can be checked against any \emph{consistent} type, $\htau'$. Fig.~\ref{fig:tconsistent} specifies the type consistency relation, written $\tconsistent{\htau}{\htau'}$, which specifies that two types are consistent if they differ only up to type holes in corresponding positions. The hole type is consistent with every type, and so, by the subsumption rule, empty expression holes can appear where an expression of any type is expected. The type consistency relation here coincides with the type consistency relation from gradual type theory by identifying the hole type with the unknown type~\cite{Siek06a}. Note that type consistency is reflexive and symmetric but it is not transitive (quite unlike subtyping, which is anti-symmetric and transitive; subtyping can be integrated into a gradual type system following \citet{Siek:2007qy}).

@@

The rule governing function application, Rule {SAp}, similarly treats expressions of hole type in function position as if they were of type $\tarr{\tehole}{\tehole}$ using the same matched arrow type judgement.

@@

Well-typed external expressions expand to well-typed internal expressions, $d$, for evaluation. The rules governing expansion are given in Fig.~\ref{fig:expansion} and the rules governing type assignment for internal expressions are given in Fig.~\ref{fig:hasType}. We will return to evaluation in Sec.~\ref{sec:evaluation}.

Expansion is specified, much like the type system for the external language, in the bidirectional style by two mutually defined judgements. The synthetic expansion judgement, $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$, synthesizes a type, $\htau$, from $\hexp$, and produces an expansion, $d$, and a hole context, $\hDelta$. We say more about hole contexts below. The analytic expansion judgement, $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$, checks $\hexp$ against $\htau$ and produces an expansion, $d$, of type $\htau'$, and a hole context, $\hDelta$. The governing theorem below establishes that the type $\htau'$ is necessarily consistent with $\htau$.
\begin{thm}[Typed Expansion]\label{thm:typed-expansion} ~
  \begin{enumerate}[nolistsep]
    \item
      If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      then $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
    \item
      If $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
      then $\tconsistent{\htau}{\htau'}$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$.
  \end{enumerate}
\end{thm}
\noindent
The reason analytic expansion produces an expansion of consistent type is because the subsumption rule, as previously discussed, allows us to check an external expression against any type consistent with the type the expression actually synthesizes, whereas every internal expression can be assigned at most one type, i.e. the following standard unicity property holds of the type assignment system: 
\begin{thm}[Type Assignment Unicity]
  If $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$
  and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$
  then $\htau=\htau'$.
\end{thm}
\noindent
Consequently, analytic expansion reports the type actually assigned to the expansion it produces. For example, we can derive that $\expandAna{\hGamma}{c}{\tehole}{c}{b}{\emptyset}$, where $\emptyset$ is the empty hole context.

@@

The hole expansion rules are the only rules that introduce hypotheses, of the form $\Dbinding{u}{\hGamma}{\htau}$, into the hole context, $\Delta$. The purpose of the hole context is to record a typing context, $\Gamma$, and a type, $\tau$, for each hole name, $u$.\footnote{We use a hole context, rather than recording the typing context and type directly on each hole closure, to ensure that all closures for a hole name have the same typing context and type.} This notation for hole contexts is taken from contextual modal type theory (CMTT) \cite{Nanevski2008}, identifying hole names with metavariables and hole contexts with metavariable contexts (we say more about the connection with CMTT below). In all four hole expansion rules, the typing context recorded in the hole context is simply the current typing context when the hole is expanded. In the synthetic hole expansion rules, {ESEHole} and {ESNEHole}, the generated hole context assigns the hole type, $\tehole$, to $u$, as in the typing rules. However, the first two premises of the expansion subsumption rule, Rule EASubsume, disallow the use of subsumption for holes in analytic position during expansion. Instead, we have separate analytic rules, {EAEHole} and {EANEHole}, which record the type that the hole is being checked against into the hole context. This is again so that we can use a type assignment system for the internal language --- the type assignment rules TAEHole and TANEHole in Fig.~\ref{fig:hasType} assign a hole closure for $u$ the corresponding type from the hole context.

@@

\subsubsection{Cast Insertion} 
Consider the following example: $\hap{(\halam{x}{\tehole}{\hap{x}{c}})}{c}$. The type synthesized for this example viewed as an external expression is $\tehole$, because the hole type annotation on $x$ allows us to apply it as a function of type $\tarr{\tehole}{\tehole}$ as previously discussed, and the outer argument, $c$, can be checked against type $\tehole$ by subsumption. However, viewed as an internal expression, this example is not well-typed---we do not have subsumption in the type assignment system defined in Fig.~\ref{fig:hasType}. Indeed, it would violate type safety if we could assign a type to this example in the internal language, because beta reduction of this example viewed as an internal expression would result in $c(c)$, which is clearly not well-typed ($c$ does not have arrow type). The problem is that leaving the argument type unknown leaves how the argument is being used (in this case, as a function) also unknown.\footnote{In a system where type reconstruction is first used to try to fill in type holes, we could express a similar example by using $x$ at two or more different types, thereby causing type reconstruction to fail.
%On the other hand, if it is acceptable to arbitrarily choose one of the possible types, and type reconstruction is complete, then type holes will never appear in the internal language and the cast insertion machinery described in this section can be omitted entirely, leaving only the hole closure machinery described previously.
}\todo{talk about elsewhere? maybe do type-hole-free version of calculus in appendix if there is time}{} Recall that we can interpret hole types as unknown types from gradual type theory. This leads us to the solution to this problem: cast insertion.

The cast form in Hazelnut Live is $\dcasttwo{\dexp}{\htau_1}{\htau_2}$, which, as specified by Rule TACast in Fig.~\ref{fig:hasType}, serves to ``box'' an expression of type $\htau_1$ for treatment as an expression of some consistent type $\htau_2$.\footnote{In the earliest work on gradual type theory, the cast form only gave the target type, $\htau_2$ \cite{Siek06a}, but it simplifies matters to include the assigned type, $\htau_1$, in the syntax \cite{DBLP:conf/snapl/SiekVCB15}.}

Casts are inserted during the expansion of function applications and ascriptions. The latter is more straightforward: Rule~{ESAsc} in Fig.~\ref{fig:expandSyn} inserts a cast from the type assigned to the ascribed expression to the ascribed type. Theorem~\ref{thm:typed-expansion} inductively ensures that the two types are consistent.  Note that we included ascription only for the sake of exposition: it can be defined using application together with the half-annotated identity function: $e : \tau = \hap{(\halam{x}{\htau}{x})}{e}$.

Cast insertion by the application expansion rule, Rule~{ESAp}, has much the same intuition, but requires a bit more care. To understand the rule, consider the expansion for the example above:
\[\hap{\dcasttwo{(\halam{x}{\tehole}{\underline{\hap{\dcasttwo{x}{\tehole}{\tarr{\tehole}{\tehole}}}{\dcasttwo{c}{b}{\tehole}}}})}{\tarr{\tehole}{\tehole}}{\tarr{\tehole}{\tehole}}}{\dcasttwo{c}{b}{\tehole}}
\]
Let us focus on the expansion of the function body, underlined\todo{shaded?}, first. A cast is inserted on both the function expression, $x$, and the the argument, $c$.

The cast on $x$ allows us to treat the variable $x$, which is of type $\tehole$, as being of the matched arrow type $\tarr{\tehole}{\tehole}$, as described in Sec.~\ref{sec:external-statics}. The first three premises of Rule~{ESAp} accomplish this by first synthesizing a type for the function expression, here $\tehole$, then determining the matched arrow type, $\tarr{\tehole}{\tehole}$, and then performing analytic expansion on the function expression with this matched arrow type. The resulting expansion will have some type $\tau_1'$ consistent with the matched arrow type. In this case, because $x$ is of variable form, analytic expansion goes through subsumption so $\tau_1'$ is simply $\tehole$. The conclusion of the rule inserts the corresponding cast. We perform type synthesis first and then use analytic expansion so that the hole context records the matched arrow type for holes in function position, rather than the type $\tehole$ for all such holes as would be the case in a variant of this rule using synthetic expansion for the function expression\todo{put this in the appendix?}{}.

The conclusion of Rule~{ESAp} also inserts a cast on the argument's expansion, from the type it assigned by the final premise of the rule, here $b$ as described at the start of Sec.~\ref{sec:expansion}, to the argument type, $\tehole$, of the matched arrow type of the function expression, here $\tehole$.

Observe that together, these casts allow us assign a type to the function body according to the rules in Fig.~\ref{fig:hasType}, where we could not do so under the same context without casts.

The outer application in the example above goes through the same rule. In this case, the cast on the function is the identity cast for $\tarr{\tehole}{\tehole}$. We do not attempt to avoid the insertion of identity casts in the core calculus for simplicity (these will simply never fail during evaluation), but in practice it is straightforward to do so (and some formal accounts of gradual typing do so by defining three application rules, including the original account of \cite{Siek06a}).

@@

Failed casts next...

\input{fig-ground-types}
\input{fig-dynamics-aux}
%% \input{fig-dynamics-steps}
\input{fig-dynamics-contexts}
